---
title: "SAIL-MENO Cohort"
output: word_document
date: "2024-11-28"
---

## Source code

Use this code to gain access to the SAIL data and create the base tables needed to create SAIL-MENO. 

You can amend this code on the source file based on your research requirements. For example, you can choose to include or exclude certain surgeries, symptoms, medications and conditions. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("0-project-setup RA.r")

source("1-base-tables.R")

```

## Menopause Demographics

Now you have all the code you need to view the menopause demographics data.

```{r demo}

meno_demographics <- runSQL(conn,paste0("select * from ",meno_demographics," ", sep=""))
head(meno_demographics)

```

## Summary Statistics Table

Create and view the summary statistics table for SAIL-MENO:

```{r create_table, echo=TRUE}

# Remove dates missing from meno_demographics:

meno_demographics <- meno_demographics%>%filter(between(FIRST_KNOWN_MENOPAUSE_START_DT,"1996-01-01", "2023-12-31"))

                              
# Get column with first menopause status 

meno_demographics <- meno_demographics%>%
  group_by(ALF_PE)%>%
  mutate(first_type=MENOPAUSE_TYPE[which.min(FIRST_KNOWN_MENOPAUSE_START_DT)])%>%
  ungroup()


# Recode all future dates to feb 2024 date (most recent verified date we have in SAIL)

meno_demographics <- meno_demographics%>%
  mutate(END_DATE=pmin(END_DATE,as.Date('2024-03-01')))

# Get durations in there 

meno_demographics <- meno_demographics%>%
  group_by(ALF_PE)%>%
  mutate(
    max_end_date=max(END_DATE, na.rm=TRUE),
    min_meno_date=min(FIRST_KNOWN_MENOPAUSE_START_DT, na.rm=TRUE),
    duration=as.numeric(max_end_date- min_meno_date))

na_count <- sum(is.infinite(meno_demographics$duration))
na_count

# We have duration rows with Inf values which means these few people had missing data for their END_DATE

# As this number is small lets impute using mean duration:

meno_demographics <- meno_demographics%>%
  group_by(ALF_PE)%>%
  mutate(
    mean_value=round(mean(duration[!is.infinite(duration)],na.rm=TRUE), 0),
    duration=ifelse(is.infinite(duration),mean_value, duration))%>%
  ungroup()%>%
  dplyr::select(-mean_value)

na_count <- sum(is.infinite(meno_demographics$duration))
na_count
  
# Create wide format data 

meno_demographics$FIRST_KNOWN_MENOPAUSE_START_DT <- as.Date(meno_demographics$FIRST_KNOWN_MENOPAUSE_START_DT)

meno_demographics$EVIDENCE_TYPE <- as.factor(meno_demographics$EVIDENCE_TYPE)

meno_demographics$MENOPAUSE_TYPE <- as.factor(meno_demographics$MENOPAUSE_TYPE)

meno_demographics[5:26] <- lapply(meno_demographics[5:26],as.character)



meno_complete <- meno_demographics%>%
  arrange(ALF_PE, FIRST_KNOWN_MENOPAUSE_START_DT)%>%
  group_by(ALF_PE)%>%
  summarise(
    MENOPAUSE_TYPE=first(MENOPAUSE_TYPE),
    duration=first(duration),
    FIRST_KNOWN_MENOPAUSE_START_DT=first(FIRST_KNOWN_MENOPAUSE_START_DT),
    EVIDENCE_TYPE=first(EVIDENCE_TYPE),
    AGE_AT_MENO=first(AGE_AT_MENO),
    TOWNSEND_2011_QUINTILE=first(TOWNSEND_2011_QUINTILE),
    across(where(is.character),~ifelse(any(.=='Y'), 'Y', 'N')))%>%
      ungroup()


# Refine surgery columns to 

# We need bilateral oophorectomy with hysterectomy column 

meno_complete$BILAT_WITH_HYST <- ifelse(meno_complete$BILATERAL_OOPHORECTOMY=='Y'& meno_complete$WITHHYSTERECTOMY=='Y', 'Y', 'N')

# We need bilateral oophprectomy only column 

meno_complete$BILAT_ONLY <- ifelse(meno_complete$BILATERAL_OOPHORECTOMY=='Y'& meno_complete$WITHHYSTERECTOMY=='N', 'Y', 'N')


# We need any HRT column 

meno_complete$HRT_REFINED <- ifelse(meno_complete$COMBINED_PATCHES=='Y'
                                    | meno_complete$COMBINED_TABLETS=='Y'
                                    | meno_complete$OESTROGEN_PATCHES=='Y'
                                    | meno_complete$OESTROGEN_TABLETS=='Y'
                                    | meno_complete$OESTROGEN_GELS=='Y'
                                    | meno_complete$OESTROGEN_IMPLANTS=='Y'
                                    | meno_complete$OESTROGEN_NASAL=='Y'
                                    | meno_complete$TESTOSTERONE =='Y'
                                    , 'Y', 'N')



## View list of variables

column_names <- names(meno_complete)
column_numbers <- seq_along(meno_complete)
column_info <- data.frame(column_name=column_names, column_number=column_numbers)
print(column_info)
    
table2 <- meno_complete[,c(2, 3, 6:7, 10, 21,  32, 13:20, 30:31)]

table2 <- table2%>%
  mutate(MENOPAUSE_TYPE=
           recode(
             MENOPAUSE_TYPE,
             Early_menopause= 'Premature Menopause',
             Surgical_menopause='Surgical Menopause',
             Symptomatic_menopause='Symptomatic Menopause'
           ))

tbl_summary(table2, 
            by=MENOPAUSE_TYPE,
            missing='no',
            type=all_continuous()~'continuous',
            statistic=all_continuous()~'{median}({p25}),({p75})',
            label=list(
              TOWNSEND_2011_QUINTILE~'Deprivation Quintile',
              duration~'Follow-up Duration (Days)',
              HOT_FLUSHES~'Hot Flushes on Record',
              HRT_REFINED~'Ever Prescribed HRT',
              COMBINED_PATCHES~'Ever Prescribed Combined Patches',
             COMBINED_TABLETS~'Ever Prescribed Combined Tablets',
              OESTROGEN_PATCHES~'Ever Prescribed Oestrogen Patches',
              OESTROGEN_GELS~'Ever Prescribed Oestrogen Gels',
              OESTROGEN_IMPLANTS~'Ever Prescribed Oestrogen Implants',
             OESTROGEN_NASAL~'Ever Prescribed Oestrogen Nasal Sprays',
              OESTROGEN_TABLETS~'Ever Prescribed Oestrogen Tablets',
              TESTOSTERONE~'Ever Prescribed Testosterone',
             BILAT_WITH_HYST~'Bilateral Oophorectomy with Hysterectomy',
              BILAT_ONLY~'Bilateral Oophorectomy without Hysterectomy',
              REPORTED_SYMPTOMS~'Menopause Symptoms on Record',
              AGE_AT_MENO~'Age at Menopause Onset'
              
            ))%>%
  bold_labels()%>%
  modify_spanning_header(all_stat_cols()~'**First Menopausal Status**')%>%
  as_gt()




```







